---
title: "Heat-related Mortality"
author: Paul Crowley
output: html_document
---

<style type="text/css">
h1.title {
font-size: 40px;
text-align: center;
}
</style>

```{r setup, include=FALSE}
source("predictgls.R")
source("stdres.gls.R")
source("moranBasis.R")
library(tidyverse)
library(sf)
library(sp)
library(spdep)
```

```{r, echo = FALSE}
heat <- st_read("HoustonHeat.shp")
head(heat)
```

## Base Linear Model

```{r, fig.align='center'}
# Fit independent MLR model with logged response
heat$log_count <- log(heat$Count + 1)
heat_lm <- lm(log_count ~ NOAC + MED_AGE + HispanicPC + BlackPCT + under5PCT + over65PCT + povertyPCT + alonePCT + MinTemp + MinTemp + Population, data = heat)

# Plot residuals
heat$residuals <- residuals(heat_lm)
ggplot(data = heat) +
  geom_sf(aes(fill = residuals), color = NA) +
  scale_fill_viridis_c() +
  labs(title = "Map of Residuals")
```

There are patches of yellow and green blocks, indicating consistent overestimation of mortality in those areas. We should test for spatial correlation.

## Moran's I Test for Spatial Correlation

```{r, echo=FALSE}
heat <- st_make_valid(heat) # Fix invalid shared vertex
```

```{r, fig.align='center'}
# Neighbors list for spatial weights
nb <- poly2nb(heat)
listw <- nb2listw(nb, style = "W")

# Perform Moran's I test
moran_result <- moran.test(residuals(heat_lm), listw)
print(moran_result)
```

The results of the Moran I test give statistically significant evidence of spatial correlation. The model must account for this correlation for accurate inference.

## Spatial Weight Matrix and Model Fit

```{r, fig.align='center'}
# Calculate Moran's Eigenvectors and create dataframe
A <- nb2mat(poly2nb(heat), style = "B")
X <- model.matrix(heat_lm)
M <- moranBasis(X, A, tol=0.95)
heatDF <- cbind(heat, M)
heatDF <- st_drop_geometry(heatDF)

# Fit GLM
heat_glm <- glm(Count ~ . - log_count - residuals, family = poisson(), data = heatDF)
```


```{r, fig.align='center'}
resid_decorrelated <- residuals(heat_glm, type = "response")
moran.test(resid_decorrelated, listw)
heat$std_resid <- scale(resid_decorrelated)
ggplot(data = heat) +
  geom_sf(aes(fill = std_resid), color = "black") +
  scale_fill_distiller(palette = "Spectral", name = "Standardized\nResiduals")
```

The residuals do not appear spatially correlated anymore based the Moran I test and this map.

## Confidence Intervals of Coefficients

```{r, fig.align='center'}
# Extract coefficients and calculate intervals
coef_summary <- summary(heat_glm)$coefficients
selected_coefs <- coef_summary[1:11, ]
ci_lower <- selected_coefs[, "Estimate"] - 1.96 * selected_coefs[, "Std. Error"]
ci_upper <- selected_coefs[, "Estimate"] + 1.96 * selected_coefs[, "Std. Error"]
conf_intervals <- data.frame(
  Lower = ci_lower,
  Upper = ci_upper
)
```
## Map Residuals

```{r, fig.align='center'}
# Plot correlated residuals
heat$dev_resid <- residuals(heat_glm, type = "deviance")
ggplot(data = heat) +
  geom_sf(aes(fill = dev_resid), color = NA) +
  scale_fill_viridis_c(option = "D") +
  labs(title = "Map of Deviance Residuals",
       fill = "Residuals") +
  theme_minimal()
```
